### 2023-03-28 definitions.tex
details

The model we are interested in is (S)WSTS (and later some particular instances, i.e. Timed/Counter Automata for instance).

@@

\begin{definition}
A {\em labeled transition system} (LTS for short) is a tuple $\mathscr{S}=(S, \Lambda, \rightarrow)$ where 
$S$ is a set of {\em states}, $ \Lambda$ is a set of {\em labels}, and 
${\rightarrow} \subseteq S\times \Lambda \times S$ is a 
ternary relation,
denoted as the set of {\em labeled transitions}. 
\end{definition}

\noindent
Labels can be used to represent the reading of an input, but also to represent an action performed during the transition or conditions that must hold in order to allow the use of the transition.

A {\em path} in a labeled transition system from a {\em source configuration} $s_0$
to a {\em target configuration} $s_n$ is a sequence 
$\pi = s_0 \xrightarrow{a_0 } s_1 \xrightarrow{a_1 } \cdots \xrightarrow{a_{n-1} } s_n$. 
We define the {\em concatenation} $ \pi_1 \pi_2$ of 
two paths $\pi_1$ and $\pi_2$ when the source configuration of $\pi_2$ is equal to the target configuration of $\pi_1$
as expected.
The {\em length} of 
$\pi = s_0 \xrightarrow{a_0 } s_1 \xrightarrow{a_1 } \cdots \xrightarrow{a_{n-1} } s_n$
is defined as $|\pi|=n$. We say the path is {\em labeled} by $a_0 a_1 , \ldots a_{n-1}$.
For all $w \in \Lambda^*$, all $s,s' \in S$, we will write $s \xrightarrow{w } s'$ if there exists a path from $s$ to $s'$ labeled by $w$.

\noindent
The set of {\em successors} of a configuration $s \in S$ is deﬁned as
$\post(s) = \{ s' \in S \mid \exists a \in \Lambda ~ s \xrightarrow{a} s'\}$.
A configuration without successors is called a {\em dead end}.  \\
The set of {\em predecessor} of a configuration $s \in S$ is deﬁned as
$\pred(s) = \{ s' \in S \mid \exists a \in \Lambda ~ s' \xrightarrow{a} s\}$.

A labeled transition system $(S, \Lambda, \rightarrow)$ is {\em deterministic} if for all configurations $s_1, s_2, s_3 \in S$ and all
$a \in \Lambda$,
 $ s_1  \xrightarrow{a} s_{2}$ and  
 $s_1  \xrightarrow{a} s_{3} $ implies $s_2 = s_3 $. \\

\begin{definition}
An {\em (unlabeled) transition system} is a pair $T = (S,\rightarrow )$ where $S$ is a set of 
{\em states} and  
$ {\rightarrow} \subseteq S \times S$ is a
binary relation 
on
the set of states, denoted as the set of {\em transitions}. 
\end{definition}

We again prefer to use infix notation and write $s \rightarrow s'$ to denote a {\em transition} from state $s$ to state $s'$ (i.e., $ (s,s') \in  {\rightarrow} $). \newline
Note 
 that an unlabeled transition system can be seen as a labeled transition system where the set of labels consists of only one element. 
Determinism, (infinite) paths, their length, and concatenation in unlabeled transition systems are
then defined as expected.

@@

@@

A {\em quasi-ordering} (a qo) is any reflexive and transitive relation $\leq$.

@@

\begin{definition}
 A {\em well-quasi-ordering} (a wqo) is any quasi-ordering $\leq$ (over some set $X$ ) such that, for any infinite sequence $x_0, x_1, x_2, ...$ in $X$, there exist indexes $i \leq j$ with
$x_i \leq  x_j$.
\end{definition}

Notice that a wqo is well-founded, i.e. it admits no infinite strictly decreasing sequence
$x_0 > x_1 > x_2 > \cdots$

% \textcolor{red}{Add lemma about infinite increasing subsequences ?}

\begin{lemma}
(Erd\"os and Rado). Assume $\leq$ is a wqo. Then any infinite sequence contains an infinite increasing subsequence: $x_{i_0} \leq x_{i_1} \leq x_{i_2} \cdots$ (with $i_0 < i_1 < i_2 \cdots$).
\end{lemma}

\iffalse
\begin{proof}
Consider an infinite sequence and the set $M = \{i \in N \mid \forall j > i ~ x_i \not\leq x_j \}$. $M$ cannot
be infinite, otherwise it would lead to an infinite subsequence contradicting the wqo
hypothesis. Thus, $M$ is bounded and any $x_i$ with $i$ beyond $M$ can start an infinite
increasing subsequence.
\end{proof}
\fi

Given $\leq$ a quasi-ordering over some set $X$, an {\em upward-closed set} is any set $I \subseteq X$ such that if $y \geq x$ and $x \in I$ then $y \in I $.
A {\em downward-closed set} is any set $I \subseteq X$ such that if $y \leq x$ and $x \in I$ then $y \in I $. 
To any $A \subseteq X$ we associate
the {\em upward-closure of $A$} 
 $\uparrow A = \{x \in X \mid \exists a \in A ~ y \geq a\}$
 and the 
 {\em downward-closure of $A$} 
 $\downarrow A = \{x \in X \mid \exists a \in A ~ y \leq a\}$. 
We abbreviate $\uparrow \{x\}$ (resp. $\downarrow \{x\}$)
as $\uparrow x$ (resp. $\downarrow x$).

A {\em basis} of an upward-closed set $I$ is a set $I_b$ such that $I = \bigcup_{x \in I_b} \uparrow x$. 
% Higman investigated ordered sets with the finite basis property.

@@

\begin{proof}
Assue we have a counter-example.
We extract an infinite subsequence where
inclusion is strict: $I_{n_0} \subsetneq I_{n_1} \subsetneq I_{n_2} \cdots$. Now, for any $i>0$, we can pick some $x_i \in I_{n_i} \setminus I_{n_{i-1}}$. The well-quasi-ordering hypothesis means that the infinite sequence of $x_i$'s
contains an increasing pair $x_i \leq x_j$ for some $i<j$. Because $x_i$ belongs to an upward-
closed set $I_{n_i}$ we have $x_j \in I_{n_i} $, contradicting $x_j \not\in I_{n_{ j - 1}}$.
\end{proof}

>>>>

The model we are interested in is WSTS (and later some particular instances, i.e. Timed/Counter Automata for instance).

@@

\begin{definition}
An {\em transition system} is a pair $\mathscr{S} = (S,\rightarrow )$ where $S$ is a set of 
{\em states} and  
$ {\rightarrow} \subseteq S \times S$ is a
binary relation 
on
the set of states, denoted as the set of {\em transitions}. A {\em labeled transition system} (LTS for short) is a tuple $\mathscr{S}=(S, \Sigma, \rightarrow)$ where 
$S$ is a set of {\em states}, $\Sigma$ is a (finite) set of {\em labels}, and 
${\rightarrow} \subseteq S\times \Sigma \times S$ is a 
ternary relation,
denoted as the set of {\em labeled transitions}. 
\end{definition}

Note 
 that an (unlabeled) transition system can be seen as a labeled transition system where the set of labels consists of only one element.

We again prefer to use infix notation and write $s \rightarrow s'$ to denote a {\em transition} from state $s$ to state $s'$ (i.e., $ (s,s') \in  {\rightarrow} $). \newline

%
\alain{enlever ce qui est inutile}
%
Determinism, (infinite) paths, their length, and concatenation in unlabeled transition systems are
then defined as expected.

@@

%
%\begin{definition}
%A {\em labeled transition system} (LTS for short) is a tuple $\mathscr{S}=(S, \Sigma, \rightarrow)$ where 
%$S$ is a set of {\em states}, $\Sigma$ is a (finite) set of {\em labels}, and 
%${\rightarrow} \subseteq S\times \Sigma \times S$ is a 
%ternary relation,
%denoted as the set of {\em labeled transitions}. 
%\end{definition}

\noindent
%Labels can be used to represent the reading of an input, but also to represent an action performed during the transition or conditions that must hold in order to allow the use of the transition.

A {\em path} in a labeled transition system from a {\em source configuration} $s_0$
to a {\em target configuration} $s_n$ is a sequence 
$\pi = s_0 \xrightarrow{a_0 } s_1 \xrightarrow{a_1 } \cdots \xrightarrow{a_{n-1} } s_n$.

We define the {\em concatenation} $ \pi_1 \pi_2$ of 
two paths $\pi_1$ and $\pi_2$ when the source configuration of $\pi_2$ is equal to the target configuration of $\pi_1$
as expected.
The {\em length} of 
$\pi = s_0 \xrightarrow{a_0 } s_1 \xrightarrow{a_1 } \cdots \xrightarrow{a_{n-1} } s_n$
is defined as $|\pi|=n$ and we say the path is {\em labeled} by $a_0 a_1 , \ldots a_{n-1}$.
For all $w \in \Sigma^*$, all $s,s' \in S$, we will write $s \xrightarrow{w } s'$ if there exists a path from $s$ to $s'$ labeled by $w$.

\noindent
The set of {\em (immediate) successors} of a state $s \in S$ is deﬁned as
$\post(s) = \{ s' \in S \mid \exists a \in \Sigma ~ s \xrightarrow{a} s'\}$.
A state without successors is called a {\em dead end}.  \\
The set of {\em (immediate) predecessor} of a state $s \in S$ is deﬁned as
$\pred(s) = \{ s' \in S \mid \exists a \in \Sigma ~ s' \xrightarrow{a} s\}$.

A labeled transition system $(S, \Sigma, \rightarrow)$ is {\em deterministic} if for all states $s_1, s_2, s_3 \in S$ and all
$a \in \Sigma$,
 $ s_1  \xrightarrow{a} s_{2}$ and  
 $s_1  \xrightarrow{a} s_{3} $ implies $s_2 = s_3 $. \\

%
\alain{enlever ce qui est inutile}
%

@@

A {\em quasi-ordering} (a qo) is any reflexive and transitive relation $\leq$ over some set $X$ and we often write $(X,\leq)$.

@@

\begin{definition}
 A {\em well-quasi-ordering} (a wqo) is any quasi-ordering $(X,\leq)$ such that, for any infinite sequence $x_0, x_1, x_2, ...$ in $X$, there exist indexes $i \leq j$ with
$x_i \leq  x_j$.
\end{definition}

There exists a stronger formulation. $(X,\leq)$ is a wqo iff any infinite sequence  $x_0, x_1, x_2, ...$ in $X$ contains an infinite increasing subsequence: $x_{i_0} \leq x_{i_1} \leq x_{i_2} \cdots$ (with $i_0 < i_1 < i_2 \cdots$).

% \textcolor{red}{Add lemma about infinite increasing subsequences ?}
%
\iffalse
%
\begin{lemma}
(Erd\"os and Rado). Assume $\leq$ is a wqo. Then any infinite sequence contains an infinite increasing subsequence: $x_{i_0} \leq x_{i_1} \leq x_{i_2} \cdots$ (with $i_0 < i_1 < i_2 \cdots$).
\end{lemma}

\begin{proof}
Consider an infinite sequence and the set $M = \{i \in N \mid \forall j > i ~ x_i \not\leq x_j \}$. $M$ cannot
be infinite, otherwise it would lead to an infinite subsequence contradicting the wqo
hypothesis. Thus, $M$ is bounded and any $x_i$ with $i$ beyond $M$ can start an infinite
increasing subsequence.
\end{proof}
%
\fi
%

Given $(X,\leq)$ a quasi-ordering, an {\em upward-closed set} is any set $U \subseteq X$ such that if $y \geq x$ and $x \in U$ then $y \in U $.
A {\em downward-closed set} is any set $D \subseteq X$ such that if $y \leq x$ and $x \in D$ then $y \in D $. 
To any subset $A \subseteq X$, we may associate
its {\em upward-closure},
 $\uparrow A = \{x \in X \mid \exists a \in A ~ y \geq a\}$
 and its 
 {\em downward-closure},
 $\downarrow A = \{x \in X \mid \exists a \in A ~ y \leq a\}$. 
We abbreviate $\uparrow \{x\}$ (resp. $\downarrow \{x\}$)
as $\uparrow x$ (resp. $\downarrow x$).

A {\em basis} of an upward-closed set $I$ is a set $I_b$ such that $I = \uparrow I_b$. 
% Higman investigated ordered sets with the finite basis property.

Wqo admits many other equivalent formulations. For instance: 
$(X,\leq)$ is a wqo if and only if any infinite increasing sequence $U_0 \subseteq U_1 \subseteq U_2 \subseteq \cdots$ of
upward-closed subsets in $X$ eventually stabilizes, 
%      i.e. there is a $k \in N$ such that $U_k = U_{k+1} = U_{k+2} = \cdots $ 
if and only if any upward-closed subset $U \subseteq X$ has a
finite basis if and only if $(X,\leq)$ is well-founded (i.e. it admits no infinite strictly decreasing sequence) and without infinite antichains (i.e. there is no infinite subset in $X$ of mutually incomparable elements).
%
\iffalse
%

@@

\begin{proof}
Assue we have a counter-example.
We extract an infinite subsequence where
inclusion is strict: $I_{n_0} \subsetneq I_{n_1} \subsetneq I_{n_2} \cdots$. Now, for any $i>0$, we can pick some $x_i \in I_{n_i} \setminus I_{n_{i-1}}$. The well-quasi-ordering hypothesis means that the infinite sequence of $x_i$'s
contains an increasing pair $x_i \leq x_j$ for some $i<j$. Because $x_i$ belongs to an upward-
closed set $I_{n_i}$ we have $x_j \in I_{n_i} $, contradicting $x_j \not\in I_{n_{ j - 1}}$.
\end{proof}
%
\fi
%
